---
title: "vehicle-colour"
author: "Michael O'Rourke"
date: "April 27, 2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
source('integrate.R')
source('functions.R')
library(magick)
knitr::opts_chunk$set(echo = TRUE)


```

##Does colour effect the value of a used car in UK auctions?

In simpler times, there were 7 colours. 7 colours that could be remembered using a simple mnemonic. These days there are nearly 2,000 pantone colours and laboratory tests estimate that humans can see approximately 10,000,000 different colours. 

Colour can influence our mood, the taste of our food, even our perception of attractiveness. And if it can influence our most primitive emotions, it should be no surprise that it can influence our buying decisions too. Marketing agencies realised this a long time ago, which is why colour psychology plays such an important part in their campaigns.

The automotive industry is not short of marketeers (or designers). Their influence can be seen in the vast array of colours available to consumers. There are literally thousands of colours available for the exterior of a car. Many with names that could be given to a pop stars first born ('Lunar sky brown' and 'Pepper dust' are a couple of my current favourites). 

Clearly, someone within the industry thinks that colour is important. But with so much variety, and so many contrasting tastes, does colour make a difference to the value of a vehicle?

Internally, our valuation experts often tell us about the importance of colour; "white cars will almost always be worth more" or "avoid dune blue coloured cars". There will always be cases where colour does have an influence, just as there will always be cases where it doesn't. What we are interested in, is whether colour has a systematic, market level effect on the value of a vehicle. All else being equal, does the colour of the vehicle influence the selling price?

###Main drivers of price

There is one thing we can be sure about, colour is not the main driver of a vehicles price. The most important factor is the type of vehicle. No one would expect the difference in price between a BMW i5 and a Nissan Micra to be the colour. Even if the BMW was in opaque couché. _https://www.digitaltrends.com/photography/pantone-448-c-ugliest-color/_
This is followed by the condition of the vehicle. A brand new, low mileage vehicle will almost always be worth more than an old, high mileage vehicle. 

Then come the market influences. Supply and demand for a particular vehicle can change over time. The hot, new hatchback can soon become garden variety through no fault of its own (here's looking at you Fiat 500). However, it is not only fashions and tastes that drive demand, market seasonality also plays its part. There is an expectation within the industry that certain vehicle types are worth more at different times of the year. For example, convertibles are thought to sell better before summer than before winter so are worth more. Likewise, 4X4s are anticipated to sell better as weather conditions worsen in late autumn.

To able to attempt to answer our question, we will need to control (as best we can) for these factors.

###Our data

The data used for this analysis is wholesale sales data from Manheim Auctions between the beginning of 2016 and the end of 2017. While this dataset is smaller than website listing data over the same period, it has a couple of advantages. The first is that it is actual transactional data. We do not need to infer a sale or be forced to assume that the selling price is the same the listing price. 

The second advantage is that there is no fixed price for a vehicle before the bidding process. Buyers base their bids based on their perception of the worth of the vehicle. This gives us a dataset that, once vehicle and market conditions have been controlled for, gives us artefacts about a vehicles relative worth. If colour is an influencer of vehicle price, we should be able to see evidence to support it.



###Controlling for main price drivers


####Difference in vehicle specification

The industry standard vehicle identifier in the market is the CAP ID/CAP Code. This identifies a vehicle to a high degree of granularity. Vehicles with the same CAP ID, have the same manufacturer, range, model, derivative, engine size, fuel type, drive train, transmission, no of doors, and many more technical specifications. They also have relatively short production life-cycles, usually between 3 to 5 years. This makes it ideal for controlling for differences in vehicle specification. There are limitations to it. Although it categorises vehicles with standard equipment (such as leather seats or navigation systems), it does not control for any optional extras a consumer might have added. These extras aren't something we can control for within our dataset.

For the purposes of this analysis, only vehicle within the same CAP ID will be compared to each other.

###Vehicle conditions 

####Age of a vehicle

With the exception of some vintage/high performance vehicles, as a car gets older it's value decreases. 

The graph below shows the relationship between age and wholesale price. We've used a random sample of the data to make it easier to see the relationship. Age, measured in months between the registration date and sold date, is along the x-axis (young vehicles towards the left, older vehicles to the right). Sold price, measured in £, is along the y-axis, lower values at the bottom, higher prices at the top. Each dot is a wholesale vehicle sale.

The red line highlights the overall relationship. There is a steep fall in price due to age in the first few months, which softens and stabilises, before gradually levelling off. This means that vehicles lose most of their value in the first couple of years, followed by a period of steady reduction in price (2-12 years), until there isn't much value left in the vehicle and the price levels off close to 0. It is easy to see why controlling for the age of a vehicle is important.

To control for the age, each vehicle will be grouped together into an age cohort. For example, vehicles aged between 12 and 18 months will be grouped together. We will only compare vehicles with the same CAP ID and the same age cohort.


```{r age and sold price relationship}
# Create a scatterplot showing age and wholesale price relationship
# Use a sample of the data so that the curve is easier to see
# Limited to £50k for visibilities sake

ggplot(sample_n(data_subset, 10000), aes(age_months, sold_price)) + geom_point(colour = 43, alpha = 0.5) + stat_smooth(method = "lm", formula = y ~ log(x), se = FALSE, colour = 'red') + labs(x = "Age (months)", y = "Sold price") + coord_cartesian(ylim = c(0,50000)) + ggtitle("Relationship between age and sold price", "Random sample of 10,000 vehicles\nUp to £50,000")


```


```{r creating age cohorts}
# Create the age cohorts
# Create a sequence of cutoffs. Use unique so that the beginning/end of each sequence is not duplicated
age_cutoffs <- unique(c(seq(3, 12, by = 3), seq(12, 48, by =6), 
                        seq(48, 144, by = 12), seq(144, 264, by = 24))) 
#Use plate instead of age

# Create labels for these cutoffs
age_labels <- as.character(age_cutoffs[1:22]) #  leave of the last cutoff as it is out of range of the data

# Use the cut function to create a new column for the age cohorts
data_subset_with_age_cohorts <- data_subset %>%
  mutate(age_cohort = cut(age_months, breaks = age_cutoffs, labels = age_labels))

```


####Mileage of a vehicle

While there is a reasonably strong correlation between the mileage of a vehicle and the age, it is important to control for mileage. A vehicle registered on the same day will age at the same rate but can have very different mileages. One could have been bought as a second car and only used for short trips around town, the other could have been bought as the primary vehicle and has a hundred mile commute to tackle each work day. 

As the graph below illustrates, the relationship between mileage and sold price is similar to the relationship between age and sold price. Controlling for mileage will be handled in the same way as age. Each vehicle will be grouped into a mileage cohort, e.g. vehicles with mileages between 10,000 and 15,000 miles will be grouped together.  

We will only compare vehicles with the same CAP ID, age cohort and mileage cohort.

```{r relationship between mileage and sold price}
# Create a scatterplot showing age and wholesale price relationship
# Use a sample of the data so that the curve is easier to see
# Limited to £50k for visibilities sake

ggplot(sample_n(data_subset, 10000), aes(mileage, sold_price)) + geom_point(colour = 43, alpha = 0.5) + stat_smooth(method = "lm", formula = y ~ log(x), se = FALSE, colour = 'red') + labs(x = "Mileage", y = "Sold price") + coord_cartesian(ylim = c(0,50000)) + ggtitle("Relationship between mileage and sold price", "Random sample of 10,000 vehicles\nUp to £50,000")
```



```{r creating mileage cohorts and seasonality}

mileage_cutoffs <- unique(c(seq(1000, 10000, by = 1000), seq(10000, 30000, by = 5000),
                          seq(30000, 100000, by = 10000), seq(100000, 260000, by = 20000)))

mileage_labels <- as.character(mileage_cutoffs[1:28])


# Create a column which month each sale occured
# This will be used to mitigate the effect of seasonality

# Create function to convert month numbers to equivalent string

convert_months <- function(x){
  ifelse(x == 1, "Jan",
  ifelse(x == 2, "Feb",
  ifelse(x == 3, "Mar",
  ifelse(x == 4, "Apr",
  ifelse(x == 5, "May",
  ifelse(x == 6, "Jun",
  ifelse(x == 7, "Jul",
  ifelse(x == 8, "Aug",
  ifelse(x == 9, "Sep",
  ifelse(x == 10, "Oct",
  ifelse(x == 11, "Nov",
  ifelse(x == 12, "Dec", NA))))))))))))
}

data_subset_cohorts <- data_subset_with_age_cohorts %>%
  mutate(mileage_cohort = cut(mileage, breaks = mileage_cutoffs, labels = mileage_labels),
         month = convert_months(month(date_sold)),
         quarter = quarter(sold_date))



```

####Condition grade

Age and mileage can be good measures of the usage of the vehicle, but they don't directly describe the physical condition of a vehicle. A well looked after, 2-year old, 20,000-mile company car is likely to be in better condition than a 2-year old, 20,000-mile family car which has been subjected to three toddlers in the back. 

Within the wholesale industry, the physical condition of a vehicle is given a grade. Within Manheim auctions, these scored range from one to five. Well-conditioned vehicles are given a grade of one, poorly conditioned vehicles are given a grade of five. Vehicles that do not meet the criteria to reach grade five are 'unclassified'. The price penalty between grade one and five can be significant. 

To control for condition we will only compare vehicles with same condition grade. Which means once we've controlled for the type and condition of a vehicle we will only be comparing vehicles with the same CAP ID, age cohort, mileage cohort and condition grade.

###Market conditions


For the purposes of this analysis, the two market conditions we will control for are inflation and seasonality.

####Seasonality 
Supply and demand can be a key influencer on the price of a vehicle at auction. If demand is high and supply is low, then the price of a vehicle can increase and when there is an oversupply of vehicles, this can reduce the price of a vehicle.

Throughout the year, supply and demand fluctuates to the tune of the market. Plate change usually preceeds an influx of part exchanges from consumers who want to upgrade to the latest and greatest models. Fleet vehicles that have finished their two year contracts come flooding back into the market. Each of these events affect the supply of these vehicles, which in turn can have an effect on the price of a vehicle. 

To fully control for these seasonal market effects is beyond the scope of this analysis. For our purposes, we will simply group together vehicles sold in witin the same quarter. Why quarters and not months? There is a delicate balance between speciticity and sample size for our vehicle groups. We are already only comparing vehicles with the same CAP ID, age cohort, mileage cohort and condition grade. Only comparing vehicles sold with the same month will create sample sizes that are too small. Also, splitting by quarters feels natural. Quarter 1 covers the post christmas buying hangover to plate change. Quarter 2 covers the effect of the plate change. Quarters 3 and 4, mirror this effect in the second half of the year. The quietish summer months and run up to the second plate change is convered by quarter 3 and Quarter 4 covers the effect of the plate change and the slow down for christmas.

```{r effect of quarters on sales/prices}

```




####Adjusting for inflation


Our data spans 24 months, over this time inflation has increased
https://www.ons.gov.uk/economy/inflationandpriceindices/articles/consumerpriceindicesabriefguide/2016


The UK Office of National Statistics (ONS) publishes monthly Consumer Price Index (CPI) figures for each major consumer sector. The closest CPI sector available for wholesale vehicles is the used car retail CPI figures. As the retail market drives the wholesale market, assuming that the wholesale CPI figures are similar to the retail CPI figures does not seem unreasonable. All sold prices will be adjusted to the December 2017 levels (the last month in the dataset).
The ONS data can be downloaded here: _link to ONS CPI scores_

```{r amend prices for inflation using cpi score}
#CPI data
#Loaded with 1,391 columns. All blank after column 9. Keep only columns I'm interested in (used car CPI)
used_car_cpi_scores <- used_car_cpi_scores %>% select(1,3)
#Rename columns
colnames(used_car_cpi_scores) <- c("title", "used_car_cpi_score")

#CPI dataset contains CPI scores at a annual, quarterly and monthly level. Use monthly
#Example format: 2016 JUL

# Create a column with the CPI data format
data_subset_cohorts_temp <- data_subset_cohorts %>%
  mutate(cpi_date_ref = convert_date_to_cpi_date(date_sold))

# Stop if any of the references are null values
stopifnot(anyNA(data_subset_cohorts_temp$date_sold)== FALSE)

# Join the CPI score data to the vehicle data using the CPI date reference
joined_data_subset <- data_subset_cohorts_temp %>%
  left_join(used_car_cpi_scores, by = c("cpi_date_ref" = "title")) %>% 
    mutate(used_car_cpi_score = as.numeric(used_car_cpi_score)) # convert CPI values to numeric

# Set December 2017 as the reference point
# All prices will be adjusted to December 2017
inflation_reference_value <- as.numeric(used_car_cpi_scores %>% filter(title == '2016 DEC') %>% select(used_car_cpi_score))

# Adjust prices
adjusted_price_dataset <- joined_data_subset %>%
  mutate(adjusted_sold_price = round((sold_price/used_car_cpi_score) * inflation_reference_value))

# Histogram showing the effect of the adjustment
# ggplot(adjusted_price_dataset %>% 
#          mutate(adjustment_percentage = ((adjusted_sold_price - sold_price)/ sold_price)), # create inflation % change
#        aes(adjustment_percentage)) + geom_histogram(fill = 43, colour = 'grey', alpha = 0.5, binwidth = 0.01) +
#   ggtitle("Distribution of inflation adjustments")

```

How does this affect the prices of our vehicles?
```{r changes in prices due to inflation}
#Compare prices that go up and prices that go down
#Plot CPI trends
#Summarise the CPI per month
monthly_cpi <- adjusted_price_dataset %>%
  mutate(month_year = round_date(date_sold, unit  = "month")) %>%
  group_by(month_year) %>%
  summarise(monthly_cpi = max(used_car_cpi_score))
#CPI trends
ggplot(monthly_cpi, aes(month_year, as.numeric(monthly_cpi))) + geom_point() + geom_line() + coord_cartesian(ylim = c(0, 100))
#Median monthly price before CPI adjustment
monthly_price_summary <- adjusted_price_dataset %>%
  mutate(month_year = round_date(date_sold, unit  = "month")) %>%
   group_by(month_year) %>%
  summarise(median_pre_adjustment_price = median(sold_price),
            median_adjusted_price = median(adjusted_sold_price))
#Plot pre adjusted prices
ggplot(monthly_price_summary, aes(month_year, median_pre_adjustment_price)) + geom_col(fill = "red", colour = "#a8aaac", alpha = 0.8) + coord_cartesian(ylim = c(0,5500)) + labs(x = "Date", y = "Median price (£)")
ggsave('../../data/vehicle-colour/pre_adjusted_col_chart.png')
#CPI trends
ggplot(monthly_cpi, aes(month_year, as.numeric(monthly_cpi))) + geom_point() + geom_line() + coord_cartesian(ylim = c(0, 100)) + labs(x = "Date", y = "CPI score")
ggsave('../../data/vehicle-colour/cpi_trends.png')
#Post adjustment
ggplot(monthly_price_summary, aes(month_year, median_adjusted_price)) + geom_col(fill = "#00528A", colour = "#a8aaac", alpha = 0.8) + coord_cartesian(ylim = c(0,5500)) + labs(x = "Date", y = "Median price (£)")
ggsave('../../data/vehicle-colour/adjusted_col_chart.png')
#save images as png
pre_adjusted_col_chart <- image_read('../../data/vehicle-colour/pre_adjusted_col_chart.png')
cpi_trends <- image_read('../../data/vehicle-colour/cpi_trends.png')
adjusted_col_chart <- image_read('../../data/vehicle-colour/adjusted_col_chart.png')

cpi_effect <- image_morph(c(pre_adjusted_col_chart, cpi_trends, adjusted_col_chart), frames = 8)
cpi_animation <- image_animate(cpi_effect, fps = 2)
print(cpi_animation)

#Save gif
#image_write(cpi_animation, "../../data/vehicle-colour/cpi_animation.gif") # this takes about 10 minutes

```


###Seasonality

Within the automotive industry, there is an assumption that certain types of vehicles have a higher demand (and therefore higher price) at different times of the year. For example, convertible vehicles are expected to be in higher demand before the summer than after. Likewise, 4x4 vehicles are expected to be in higher demand in the run up to winter.



_Something about plate change?_

To take this into effect, sales will be split into cohorts based on the time of the year of their sale. If our dataset was larger, the cohorts could be as granular as months. Unfortunately, when you take into consideration the vehicle type, age, mileage and condition cohorts the number of vehicles available for comparison is likely to be small. For that reason, the cohorts will be at quarter level.

```{r create seasonality cohorts}

seasonality_dataset <- adjusted_price_dataset %>%
  mutate(seasonality_cohort =  quarter(date_sold))

```

###How does demand change throughout the year?

```{r how does demand change throughout the year?}
#Create data model with total sales per month per body style
#Do certain body styles have seasonal demand
seasonal_demand <- adjusted_price_dataset %>%
  #Create month variable
  mutate(month_of_year = format(date_sold,"%B")) %>%
  #consolidate body styles
  mutate(body_style = case_when(body_style == "Cabriolet" ~ "Convertible",
                                body_style == "Roadster" ~ "Convertible",
                                body_style == "Soft-Top" ~ "Convertible",
                                body_style == "Hardtop" ~ "Convertible",
                                body_style == "Tourer" ~ "Estate",
                                TRUE ~ body_style)) %>%
  #Count number of sales per body style per month of the year
  group_by(body_style, month_of_year) %>%
  summarise(no_of_sales = n()) 
  #convert and reorder month_of_year
  seasonal_demand$month_of_year <- factor(seasonal_demand$month_of_year, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
  
#Total sales per body style
total_sales_per_body_style <- adjusted_price_dataset %>%
  group_by(body_style) %>%
  summarise(total_sales = n())

#Join back seasonal demand

seasonal_demand_enhanced <- seasonal_demand %>%
  left_join(total_sales_per_body_style, by = "body_style") %>%
  mutate(percent_of_total = no_of_sales/total_sales)

#Plot sales per month
#Do not have values on the y-axis. Sensitive information
ggplot(seasonal_demand, aes(x = month_of_year, y = no_of_sales)) + geom_col(fill = "#00528A", colour = "#a8aaac", alpha = 0.8) + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + labs(x = "Month", y = "Sales")




img <- image_graph(600, 340, res = 96)
datalist <- split(seasonal_demand_enhanced, seasonal_demand_enhanced$body_style)
out <- lapply(datalist, function(data_){
  p <- ggplot(data_, aes(x = month_of_year, y = percent_of_total)) + geom_col(fill = "#00528A", colour = "#a8aaac", alpha = 0.8) + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + labs(x = "Month", y = "% of total sales") + ggtitle(data_$body_style)
  print(p)
})
dev.off()
sales_per_body_style <- image_animate(img, fps = 0.5)
print(sales_per_body_style)
image_write(sales_per_body_style, "../../data/vehicle-colour/sales_per_body_style.gif") 

```



To summarise, only vehicles with the same cap ID, age cohort, mileage cohort, condition grade and seasonality cohort will be compared. In addition, all prices have been adjusted for inflation to December 2017 prices. 

That's a lot of restrictions, will there be any groups of vehicles with large enough sample sizes to compare against each other?

```{r sample size of the cohorts}
# Create groups with the same age, mileage, condition, cap_id, and seasonality
vehicle_cohort_dataset <- seasonality_dataset %>%
  mutate(vehicle_group = paste(cap_id, age_cohort, mileage_cohort, condition_grade, seasonality_cohort, sep = '-'))

# count how many observations per vehicle group
observations_per_vehicle_group <- vehicle_cohort_dataset %>%
  group_by(vehicle_group) %>%
  summarise(no_of_observations = n()) %>%
  arrange(-no_of_observations) 

ggplot(observations_per_vehicle_group %>% top_n(20, no_of_observations), 
       aes(reorder(vehicle_group, -no_of_observations), no_of_observations)) +
  geom_col(fill = "#00528A", colour = "#a8aaac", alpha = 0.8) + labs(x = "Vehicle group", y = "No of observations") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("Number of observations per vehicle group", "Top 20 vehicle groups")

```

As expected, the sample size of each group is quite small. The largest group has only 113 observations, the 10th has 49 and the 20th has 34. As much as we would like these groups to be larger, the importance of specificity far outweighs the need for large sample sizes.

####What should be our minimum sample size?

If our cut off is too high, then there are two problems. The most obvious is that it gives us too few groups to compare colours in. The second, is to do with the variety of vehicles. The eagle eyed amongst us would have noticed that the CAP IDs of the top 20 groups are very similar. In fact, 14 of the top 20 groups are Fiesta's, and 17 are Fords. If we set the cut off too high, we will only be able to compare how colour affects young, low mileage, Ford Fiestas.

If our cut off it too low, then we end up with thousands of groups with a couple of vehicles to compare which is unlikely to tell us anything more than random chance would. Approximately, 99% of all the vehicle groups have 5 or less observations.

Ten observations feels too low and twenty observations reduces the breadth of vehicle models we can compare. A reasonable compromise seems _why is this reasonable_ to be 15 observations per vehicle group. This gives us 232 vehicle groups to compare.

###Colours

The simple days of only having to remember the seven colours of the rainbow have long done. In a world where there are nearly 2,000 pantone colours available and even black has different shades. It should be no surprise that there are 1,614 'different' colours of vehicles within our dataset. When designers and marketing teams combine the possibilities are endless. A couple favourites are 'Lunar sky brown' and 'Pepper dust'.

For this analysis, we'll be going back to basics with our colour descriptions. 'Moroccan red' and 'Tango red metallic' will be both classified as 'red', 'Sea grey' and 'Storm grey' are both 'grey'. This will result in some dropped observations, what colour of the rainbow is 'fashionista' or 'vision'.

```{r consolidating colours}
list_of_colour <- as.data.frame(cbind(exterior_body_colour = unique(seasonality_dataset$exterior_body_colour)))

consolidate_colour <- function(x){
  x %>% mutate(colour_group = case_when(str_detect(exterior_body_colour, "GREEN") ~ "green",
                         str_detect(exterior_body_colour, "SILVER") ~ "silver",
                         str_detect(exterior_body_colour, "BLUE") ~ "blue",
                         str_detect(exterior_body_colour, "GREY") ~ "grey",
                         str_detect(exterior_body_colour, "GOLD") ~ "gold",
                         str_detect(exterior_body_colour, "ORANGE") ~ "orange",
                         str_detect(exterior_body_colour, "TURQUOISE") ~ "green",
                         str_detect(exterior_body_colour, "BLACK") ~ "black",
                         str_detect(exterior_body_colour, "BRONZE") ~ "orange",
                         str_detect(exterior_body_colour, "RED") ~ "red",
                         str_detect(exterior_body_colour, "WHITE") ~ "white",
                         str_detect(exterior_body_colour, "YELLOW") ~ "yellow",
                         str_detect(exterior_body_colour, "SAPPHIRE") ~ "blue",
                         str_detect(exterior_body_colour, "GREEN") ~ "green",
                         # str_detect(exterior_body_colour, "GREEN") ~ "green",
                         # str_detect(exterior_body_colour, "GREEN") ~ "green",
                         # str_detect(exterior_body_colour, "GREEN") ~ "green",
                         # str_detect(exterior_body_colour, "GREEN") ~ "green",
                         TRUE ~ "other"
  ))
}

colour_temp <- consolidate_colour(vehicle_cohort_dataset)

#Plot the number of observations per colour
ggplot(colour_temp %>% group_by(colour_group) %>% summarise(n = n()), 
       aes(reorder(colour_group, -n), n)) +
  geom_col(fill = "#00528A", colour = "#a8aaac", alpha = 0.8) + labs(x = "Colour", y = "No of observations") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("Number of observations per colour")

#Show the reduction in colours
# non_grouped_colour_heatmap <- ggplot(colour_temp %>% group_by(exterior_body_colour) %>% summarise(median_sold_price = median(sold_price), no_of_vehicles = n()), aes(x= exterior_body_colour, y = median_sold_price)) + geom_raster(aes(fill = no_of_vehicles), interpolate = T)

```


```{r}
#Only include vehicle groups with more than n observations
n <- 10
vehicle_group_above_min_observations <- observations_per_vehicle_group %>%
  filter(no_of_observations >= n)

vehicle_group_subset <- colour_temp %>%
  filter(vehicle_group %in% vehicle_group_above_min_observations$vehicle_group)

#Rank the colours based on the median sold price
colour_per_vehicle_group <- vehicle_group_subset %>%
  group_by(vehicle_group, colour_group) %>%
  summarise(no_of_vehicles = n(),
            median_sold_price = median(sold_price)) %>%
  mutate(sold_price_rank = order(vehicle_group, colour_group, decreasing = T))

#boxplot of the distribution of ranks
ggplot(colour_per_vehicle_group, aes(x = as.factor(colour_group), y = sold_price_rank)) + geom_boxplot(colour = 43) + geom_jitter(alpha = 0.2) + scale_y_reverse() + labs(x = "Colour", y = "Median price rank")

#Need to only include vehicle groups with a variety of colours
#Only include vehicle groups with more than n colours
colour_variety_per_vehicle_group <- colour_per_vehicle_group %>%
  group_by(vehicle_group) %>%
  tally() %>%
  filter(n>=4)

#boxplot of the distribution of ranks
ggplot(colour_per_vehicle_group %>% filter(vehicle_group %in% colour_variety_per_vehicle_group$vehicle_group), aes(x = as.factor(colour_group), y = sold_price_rank)) + geom_boxplot(colour = 43) + geom_jitter(alpha = 0.2) + scale_y_reverse() + labs(x = "Colour", y = "Median price rank")

```



```{r should I use log(sold)price instrad of price}
#What is the distribution of sold prices?

ggplot(seasonality_dataset, aes(sold_price)) + geom_histogram(binwidth = 500) + geom_vline( xintercept = median(seasonality_dataset$sold_price)) + geom_vline( xintercept = mean(seasonality_dataset$sold_price), colour = "red")

ggplot(seasonality_dataset, aes(log(sold_price))) + geom_histogram(binwidth = 0.5) + geom_vline( xintercept = median(log(seasonality_dataset$sold_price))) + geom_vline( xintercept = mean(log(seasonality_dataset$sold_price)), colour = "red")


```


Test statistics for the median
moods median test pearsons chi squared test
Potentially use old school test stats
Matching inference (not for now)




































